Data Cleaning for Raw Tick-Bourne Cases Data - 2019 to 2024

### 2025-04-21 code/analysis-code/statistical-analysis.qmd
Part 4/5
Catching Up - Part Four & Part Five, citations still need to be figured out and some more fun graphics (my findings are boring) but closer?

# Save plot
plot_file1 <- here::here("results", "figures", "trend_over_time_by_disease.png")
ggsave(plot_file1, plot = plot1, width = 8, height = 6, dpi = 300)
```

```{r}
# 2. Distribution Across Health Districts:
# Bar plot of cases by health district
plot2 <- ggplot(cleaned_data3, aes(x = COUNTY, fill = DISEASE)) +
  geom_bar() +
  labs(
    title = "Cases by Health District",
    x = "Health District",
    y = "Number of Cases"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

# Save plot
plot_file2 <- here::here("results", "figures", "cases_by_health_district.png")
ggsave(plot_file2, plot = plot2, width = 8, height = 6, dpi = 300)
```

@@

# Save the plot to the results/figures folder
plot_file3 <- here::here("results", "figures", "completeness_over_time.png")
ggsave(plot_file3, plot = plot3, width = 8, height = 6, dpi = 300)

@@

# Save plot
plot_file4 <- here::here("results", "figures", "days_to_lab_over_time.png")
ggsave(plot_file4, plot = plot4, width = 8, height = 6, dpi = 300)

@@

# Save plot
plot_file5 <- here::here("results", "figures", "relative_incidence_2019_2024.png")
ggsave(plot_file5, plot = plot5, width = 8, height = 6, dpi = 300)
```

>>>>

# Save plot
plot_file1 <- here::here("results", "supplemental-figures", "trend_over_time_by_disease.png")
ggsave(plot_file1, plot = plot1, width = 8, height = 6, dpi = 300)
```

```{r}
##Distribution Across Counties:
# Bar plot of cases by county
plot2 <- ggplot(cleaned_data3, aes(x = COUNTY, fill = DISEASE)) +
  geom_bar() +
  labs(
    title = "Cases by County",
    x = "County",
    y = "Number of Cases"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

# Save plot
plot_file2 <- here::here("results", "supplemental-figures", "cases_by_eachcounty.png")
ggsave(plot_file2, plot = plot2, width = 8, height = 6, dpi = 300)
```

@@

# Save the plot 
plot_file3 <- here::here("results", "supplemental-figures", "completeness_over_time.png")
ggsave(plot_file3, plot = plot3, width = 8, height = 6, dpi = 300)

@@

# Save plot
plot_file4 <- here::here("results", "supplemental-figures", "days_to_lab_over_time.png")
ggsave(plot_file4, plot = plot4, width = 8, height = 6, dpi = 300)

@@

# Save plot
plot_file5 <- here::here("results", "supplemental-figures", "relative_incidence_2019_2024.png")
ggsave(plot_file5, plot = plot5, width = 8, height = 6, dpi = 300)
```

```{r}
## ---- modeling_setup --------
library(tidymodels)

# Load the processed dataset
model_data <- readr::read_csv(here::here("data", "processed-data", "time_between_processeddata.csv"))

# Add the target variable and convert appropriate columns
model_data <- model_data %>%
  mutate(
    delay_over_threshold = if_else(TOTAL_DAYS > 30, 1, 0),  # Create binary target
    delay_over_threshold = as.factor(delay_over_threshold),  # Convert to factor
    DISEASE = as.factor(DISEASE),
    COUNTY = as.factor(COUNTY),
    SEX = as.factor(GENDER),
    AGE = as.numeric(AGE)  # Ensure numeric
  ) %>%
  drop_na(delay_over_threshold, AGE, SEX, COUNTY, DISEASE)  # Clean up

# Save the updated dataset with the new variable
write_csv(model_data, here::here("data", "processed-data", "time_between_processeddata.csv"))

To explore whether certain demographic or clinical variables could predict delayed follow-up in confirmed tick-borne disease cases, we created a binary outcome variable called delay_over_threshold, coded as 1 if the total time between disease onset and case closure exceeded 30 days, and 0 otherwise. This variable was calculated based on the sum of time intervals between key date fields (DOO, LABDATE1, LABDATE2, and UPDATEDATE), which had been previously cleaned and processed. Predictors selected for this analysis included age, sex, county, and disease type, as these are commonly available surveillance variables that may relate to access to care or follow-up procedures. All categorical predictors were converted into factor variables to prepare them for modeling.

```{r}
## ---- modeling_setup --------
library(tidymodels)

# Load the processed dataset
model_data <- readr::read_csv(here::here("data", "processed-data", "time_between_processeddata.csv"))

# Convert appropriate variables to factors
model_data <- model_data %>%
  mutate(
    delay_over_threshold = as.factor(delay_over_threshold),  # target variable
    DISEASE = as.factor(DISEASE),
    COUNTY = as.factor(COUNTY),
    SEX = as.factor(GENDER),
    AGE = as.numeric(AGE)  # just in case it's not numeric yet
  ) %>%
  drop_na(delay_over_threshold, AGE, SEX, COUNTY, DISEASE)  # Remove any rows with missing predictors

The data was randomly partitioned into training and testing sets using an 80/20 split, stratified by the outcome variable to maintain a balanced class distribution. The training set was used for model training and cross-validation, while the test set was reserved for final evaluation of model performance on unseen data.

```{r}
set.seed(123)
data_split <- initial_split(model_data, prop = 0.8, strata = delay_over_threshold)
train_data <- training(data_split)
test_data <- testing(data_split)

A recipe() was used to define preprocessing steps applied uniformly across all models. Categorical variables (such as county, disease, and sex) were one-hot encoded using dummy variables. Zero-variance predictors were removed, and the single numeric predictor, age, was normalized to ensure comparability across different model types.
```{r}
delay_recipe <- recipe(delay_over_threshold ~ AGE + SEX + COUNTY + DISEASE, data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%  # convert categorical predictors to dummy variables
  step_zv(all_predictors()) %>%  # remove zero-variance predictors
  step_normalize(all_numeric_predictors())  # standardize age
```

```{r}
logistic_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

tree_model <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")

rf_model <- rand_forest(trees = 500) %>%
  set_engine("ranger") %>%
  set_mode("classification")

For each model, a workflow was constructed using the tidymodels framework to combine the preprocessing recipe and the corresponding model specification. This approach ensured consistent preprocessing and simplified model training and evaluation.
```{r}
logistic_workflow <- workflow() %>%
  add_model(logistic_model) %>%
  add_recipe(delay_recipe)

tree_workflow <- workflow() %>%
  add_model(tree_model) %>%
  add_recipe(delay_recipe)

rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(delay_recipe)
```
To evaluate model performance and avoid overfitting, we implemented 5-fold cross-validation repeated 3 times. This resampling strategy provides a robust estimate of out-of-sample performance. Each model was trained and evaluated on multiple folds using two performance metrics: accuracy and the area under the ROC curve (AUC), the latter of which is especially informative for imbalanced or binary classification problems.
```{r}
set.seed(123)
cv_folds <- vfold_cv(train_data, v = 5, repeats = 3, strata = delay_over_threshold)

# Fit all models
logistic_res <- fit_resamples(logistic_workflow, resamples = cv_folds, metrics = metric_set(roc_auc, accuracy))
tree_res     <- fit_resamples(tree_workflow, resamples = cv_folds, metrics = metric_set(roc_auc, accuracy))
rf_res       <- fit_resamples(rf_workflow, resamples = cv_folds, metrics = metric_set(roc_auc, accuracy))

```
Cross-validation results for each model were collected and compared using mean accuracy and AUC. These metrics provided insight into the predictive quality of each approach. The model with the best combination of high accuracy and AUC was selected for final evaluation on the test set.
```{r}
collect_metrics(logistic_res)
collect_metrics(tree_res)
collect_metrics(rf_res)
```

```{r}
## ---- final_rf_evaluation --------
# Fit the final random forest model on training data and evaluate on test data
final_rf_fit <- rf_workflow %>%
  last_fit(split = data_split)

# Collect performance metrics
collect_metrics(final_rf_fit)

# Confusion matrix
collect_predictions(final_rf_fit) %>%
  conf_mat(truth = delay_over_threshold, estimate = .pred_class)

# ROC curve
collect_predictions(final_rf_fit) %>%
  roc_curve(truth = delay_over_threshold, .pred_1) %>%
  autoplot()

# Save ROC curve plot
roc_plot <- collect_predictions(final_rf_fit) %>%
  roc_curve(truth = delay_over_threshold, .pred_1) %>%
  autoplot() +
  theme_minimal() +
  labs(title = "ROC Curve for Random Forest Model")

ggsave(
  filename = here::here("results", "supplemental-figures", "roc_curve.png"),
  plot = roc_plot,
  width = 6, height = 5, dpi = 300
)
```

### 2025-04-21 code/processing-code/processing-code-tickcases.qmd
Part 4/5
Catching Up - Part Four & Part Five, citations still need to be figured out and some more fun graphics (my findings are boring) but closer?

# Create histogram of disease counts
disease_plot <- ggplot(cleaned_data3, aes(x = DISEASE)) +
  geom_bar(fill = "#2C3E50", color = "white") +  # Dark blue bars with white edges
  theme_minimal() +
  labs(
    title = "Distribution of Tick-Bourne Disease\nTypes for Confirmed Cases in Georgia (2019-2024)",
    x = "Disease Type",
    y = "Count"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate & reduce x-axis label size
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.margin = margin(10, 10, 20, 10)  # Adjust margins to prevent cutoff
  )

@@

# Define save locations
save_plot_location <- here::here("results", "figures", "disease_histogram.png")

# Ensure directory exists
dir.create(here::here("results", "figures"), recursive = TRUE, showWarnings = FALSE)

# Save the plot
ggsave(filename = save_plot_location, plot = disease_plot, width = 12, height = 6, dpi = 300)
```

@@

# Define save location
save_rmsf_plot_location <- here::here("results", "figures", "rmsf_titer_boxplot_log.png")

@@

# Define save location
save_top5_disease_plot <- here::here("results", "figures", "top5_diseases_plot.png")

@@

# Define save location
save_lyme_count_plot <- here::here("results", "figures", "lyme_result_count_histogram2.png")

@@

# Define save location
save_disease_county_plot <- here::here("results", "figures", "disease_county_plot.png")

@@

# Define save location
save_top5_stacked_plot <- here::here("results", "figures", "top5_diseases_stacked_plot.png")

@@

# Save the plot to results/figures using here::here()
ggsave(
  filename = here::here("results", "figures", "disease_count_by_age.png"), 
  plot = age_group_plot, 
  width = 8, 
  height = 6, 
  dpi = 300
)

@@

# Save the plot to results/figures using here::here()
ggsave(
  filename = here::here("results", "figures", "status_completeness_plot.png"), 
  plot = status_completeness_plot, 
  width = 8, 
  height = 6, 
  dpi = 300
)
status_completeness_plot

@@

>>>>

# Create histogram of disease counts
disease_histogram <- ggplot(cleaned_data3, aes(x = DISEASE)) +
  geom_bar(fill = "#2C3E50", color = "white") +  # Dark blue bars with white edges
  theme_minimal() +
  labs(
    title = "Distribution of Tick-Bourne Disease\nTypes for Confirmed Cases in Georgia (2019-2024)",
    x = "Disease Type",
    y = "Count"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate & reduce x-axis label size
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.margin = margin(10, 10, 20, 10)  # Adjust margins to prevent cutoff
  )

@@

# Define save locations
save_plot_location <- here::here("results", "supplemental-figures", "disease_histogram.png")

# Ensure directory exists
dir.create(here::here("results", "supplemental-figures"), recursive = TRUE, showWarnings = FALSE)

# Save the plot
ggsave(filename = save_plot_location, plot = disease_histogram, width = 12, height = 6, dpi = 300)
```

@@

# Define save location
save_rmsf_plot_location <- here::here("results", "supplemental-figures", "rmsf_titer_boxplot_log.png")

@@

# Define save location
save_top5_disease_plot <- here::here("results", "supplemental-figures", "top5_diseases_plot.png")

@@

# Define save location
save_lyme_count_plot <- here::here("results", "supplemental-figures", "lyme_result_count_histogram2.png")

@@

# Define save location
save_disease_county_plot <- here::here("results", "supplemental-figures", "disease_county_plot.png")

@@

# Define save location
save_top5_stacked_plot <- here::here("results", "supplemental-figures", "top5_diseases_stacked_plot.png")

@@

# Save the plot to results/supplemental-figures using here::here()
ggsave(
  filename = here::here("results", "supplemental-figures", "disease_count_by_age.png"), 
  plot = age_group_plot, 
  width = 8, 
  height = 6, 
  dpi = 300
)

@@

# Save the plot to results/supplemental-figures using here::here()
ggsave(
  filename = here::here("results", "supplemental-figures", "status_completeness_plot.png"), 
  plot = status_completeness_plot, 
  width = 8, 
  height = 6, 
  dpi = 300
)
status_completeness_plot

@@

# Create binary target variable for modeling: delay > 30 days = 1, else 0
processed_time_data2 <- processed_time_data2 %>%
  mutate(
    delay1 = as.numeric(LABDATE1 - DOO),
    delay2 = as.numeric(LABDATE2 - LABDATE1),
    delay3 = as.numeric(UPDATEDATE - LABDATE2),
    total_delay_days = delay1 + delay2 + delay3,
    delay_over_threshold = if_else(total_delay_days > 30, 1, 0)
  )
